---
output:
  md_document:
    variant: gfm
---

```{r include = FALSE}
devtools::load_all("../sdmEvalToolCore")
```

# SDM Model Evaluation Tool User Guide for Modelers

> How to manage materials and deployments

## Installation

Install the following R packages:

- sf
- terra
- DBI
- RSQLite
- jsonlite
- suntools
- sdmEvalToolCore
- sdmEvalToolUI

## Uploading model materials

First, download the example model materials [BAM demo.zip](https://drive.google.com/file/d/1CJZ_TDy5XPF3UtyDz6IrWzuShyO1bdxx/view?usp=drive_link) to your sdmEvaluationTool project folder. 

Open the [`data-loading-example_JH.R`](data-loading-example_JH.R) script.

The script begins by loading the required libraries, unziping the demo 
materials (if applicable), removing old prepared materials (if applicable),
and setting up variables:

```R
library(sf)
library(terra)
library(mefa4)
library(DBI)
library(RSQLite)
library(jsonlite)
library(suntools)
library(sdmEvalToolCore)
library(sdmEvalToolUI)

path <- "./BAM Demo"
if(!exists(path)){
  if(!file.exists("./BAM Demo.zip")){
    stop("Missing BAM Demo.zip model materials. Download from https://drive.google.com/file/d/1CJZ_TDy5XPF3UtyDz6IrWzuShyO1bdxx/view?usp=drive_link to your project folder.")
  }
  unzip("./BAM Demo.zip")
}

conf <- yaml::read_yaml(system.file("config.yml", package = "sdmEvalToolCore"))
DIR <- "misc/sdm_evaluation_results"
dir.create(DIR, recursive = TRUE)

sdmevaltool_options(base = DIR)

unlink(DIR, recursive = TRUE)
dir.create(DIR, recursive = TRUE)
```

Next, create a new database file and turn foreign key constraints on:

```R
con <- db_connect(make_target_path("sdm_evaluation_db.sqlite"))
rs <- DBI::dbSendQuery(con, "PRAGMA foreign_keys = ON;")
DBI::dbClearResult(rs)
```

Use the `db_create_tables()` function to create the empty tables:

```R
db_create_tables(con)
```

Find an overview of the database structure in the system [`spec`](./spec/).
Instead of creating a new database, if you have an existing database 
file, you can copy it to the root of the `DIR` folder.

Due to the foreign key constraint, tables need to be updated respecting the
keys relationships. If a key is used as a foreign key, it needs to exist in
the parent table as a primary key.

The components table should not change, the defaults are used from the package:
```R
components <- sdmEvalToolCore::components[, c(
  "component",
  "description",
  "mandatory"
)]
colnames(components) <- c(
  "component_id",
  "component_description",
  "component_mandatory"
)
db_write_table(con, "components", components)
```

The species table should also not change frequently. However, for 
the purpose of this example, here only a small portion of all 
expected species are listed:

```R
species <- structure(
  list(
    species_id = c(
      "BLPW",
      ...,
      "TEWA"
    ),
    scientific_name = c(
      "Setophaga striata",
      ...,
      "Oreothlypis peregrina"
    ),
    english_name = c(
      "Blackpoll Warbler",
      ...,
      "Tennessee Warbler"
    ),
    french_name = c(
      "Paruline rayée",
      ...,
      "Paruline obscure"
    )
  ),
  row.names = c(24L, 40L, 76L, 90L, 106L, 118L),
  class = "data.frame"
)
db_write_table(con, "species", species)
```

Next, build the model table:

```R
model_id <- "bam_v5_can71" # we'll use the model ID later

models <- data.frame(
  model_id = model_id,
  model_name = "BAM v5 Can 71",
  model_description = "BAM version 5 Canada model in BCR 71"
)
db_write_table(con, "models", models)
```

Create the users table next:

```R
user_id <- "testuser" # user who uploads the materials

users <- data.frame(
  user_id = "testuser",
  user_name = "Test User",
  user_email = "x@y.z",
  user_affiliation = "XYZ",
  admin = FALSE,
  password = generate_password()
)
db_write_table(con, "users", users)
```

Add as many users as needed. You can make a csv file following the column names.
Read in the file with `read.csv()` to have the full `users` table.

Next, the script will load existing files containing the data for 
each model material and prepare it in the required format for the 
app. You can inspect each file for a reference of how they 
should be provided.

First, create the predictor metadata:

```R
rule <- get_comp_rule("predictor_metadata", "upload")
fi <- file.path(path, "predictors", "predictor_metadata.csv")
x <- read_file(fi)
names(x)[names(x)=="variable"]="predictor"
x <- x[, rule$output$columns]

prep_predictor_metadata(
  x = x,
  model_id = model_id,
  user_id = user_id,
  material_settings = NULL,
  con = con,
  update = FALSE
)
```

Create the model metadata:

```R
fi <- file.path(path, "metadata", "metadata_Knight_2025-12-16.csv")
x <- read_file(fi)

prep_model_metadata(
  x = x,
  model_id = model_id,
  user_id = user_id,
  material_settings = NULL,
  con = con,
  update = FALSE
)
```

To prepare the predictor raster (resampled to save space), run:

```R
fi <- file.path(path, "predictors", "predictor_stack.tif")
x <- read_file(fi)

prep_predictor_raster(
  x = x,
  resolution = 5,
  model_id = model_id,
  user_id = user_id,
  material_settings = NULL,
  con = con,
  update = FALSE
)
```

To prepare the species observations, processing multiple species at once, run:

```R
fi <- file.path(path, "data", "observations_can71.csv")
x <- read_file(fi)
colnames(x)[1:7] <- c("id", "latitude", "longitude",
  "buffer", "time", "method", "year")
prep_observations(
  x = x,
  species = species,
  model_id = model_id,
  user_id = user_id,
  material_settings = NULL,
  con = con,
  update = FALSE
)
```

The model summaries for the species are prepared with:

```R
fi <- file.path(path, "predictors", "predictor_importance.csv")
x <- read_file(fi)
colnames(x)[1] <- "species_id"
x$bcr <- NULL

prep_model_summary(
  x = x,
  species = species,
  model_id = model_id,
  user_id = user_id,
  material_settings = NULL,
  con = con,
  update = FALSE
)
```

The model fit for the species are prepared with:

```R
fi <- file.path(path, "reliability", "validation_can71.csv")
x <- read_file(fi)
colnames(x)[1] <- "species_id"
x$scientific <- x$english <- x$region <- NULL

prep_model_fit(
  x = x,
  species = species,
  model_id = model_id,
  user_id = user_id,
  material_settings = NULL,
  con = con,
  update = FALSE
)
```

Lastly, the species predictions are prepared with:

```R
for (species_id in species$species_id) {
  fi <- file.path(path, "predictions", paste0(species_id, "_can71_2020.tif"))
  x <- read_file(fi)

  prep_spatial_prediction(
    x = x,
    model_id = model_id,
    species_id = species_id,
    user_id = user_id,
    material_settings = NULL,
    con = con,
    update = FALSE
  )
}
```

## Material settings

Default material settings are defined in the YAML specification.
To access these, you can run the following directly in the RStudio Console:

```{r}
component_id <- "predictor_metadata"
str(get_comp_rule(component_id, "display")$material_settings)
```

This default is used when using the above `prep_<component_id>()` functions.
Using the same structure, we can pass a new legend:

```R

fi <- file.path(path, "predictors", "predictor_metadata.csv")
x <- read_file(fi)
names(x)[names(x)=="variable"]="predictor"
x <- x[, rule$output$columns]

prep_predictor_metadata(
  x = x,
  model_id = model_id,
  user_id = user_id,
  material_settings = list(en = "Table with predictor metadata.", fr = ""),
  con = con,
  update = TRUE
)
```

Using `update = TRUE` we can update an existing material.

## Managing deployments

Create 2 deployments, keeping the `deploylemt_settings` empty (`[]`) for now.
This field is JSON, we'll add the value later:

```R
deployments <- data.frame(
  deployment_id = c("deployment1", "deployment2"),
  deployment_name = c("Complex for spatial prioritization", "Simple for population assessment"),
  deployment_description = c(
    "Complex for spatial prioritization",
    "Simple for population assessment"
  ),
  deployment_create_user = "testuser",
  deployment_create_time = timestamp_to(now()),
  deployment_settings = c("[]", "[]")
)
```

Let's read in the welcome message (markdown file):

```R
rule <- get_comp_rule("deployment_settings", "upload")
fi <- file.path(path, "welcome", "evaluator_welcome.md")
x <- read_file(fi)
```

Next we define the `deployment_settings` using the welcome message
and the use case:

```R
d1 <- d2 <- conf$templates$deployment_settings
d1$use_case <- list(list(en = "Spatial prioritization", fr = ""))
d2$use_case <- list(list(en = "Population assessment", fr = ""))
d1$instructions_to_evaluators <- paste0(x, collapse = "\n")
d2$instructions_to_evaluators <- paste0(x, collapse = "\n")
deployments$deployment_settings[1] <- jsonlite::toJSON(d1)
deployments$deployment_settings[2] <- jsonlite::toJSON(d2)
```

We write the table to the database:

```R
db_write_table(con, "deployments", deployments)
```

Write the deployment settings to the files as well:

```R
for (j in 1:nrow(deployments)) {
  prep_deployment_settings(
    deployment_id = deployments$deployment_id[j],
    x = fromJSON(deployments$deployment_settings[j])
  )
}
```

The default deployment questions can be written to the files using the following
expression (you can also provide your own as the argument `x`):

```R
prep_deployment_questions(
  deployment_id = "deployment1",
  x = NULL
)
```

The ./misc/sdm_evaluation_results/deployments/deployment1/deployment_questions.csv file now can be edited as needed.

The question set can also be edited before writing the csv file. In this example we simplify the evaluation by removing the followup, model_fit and model_summary questions. 

```R
q <- sdmEvalToolCore::default_questions
q$followup_level <- 0
q <- subset(q,!is.element(component,c("model_fit","model_summary")))
q <- combine_questions(q)

prep_deployment_questions(
  deployment_id = "deployment2",
  x = q
)
```

Define user access to the deployments (add evaluator user names here):

```R
access <- data.frame(
  user_id = user_id,
  deployment_id = deployments$deployment_id,
  user_roles = "evaluator"
)
db_write_table(con, "access", access)
```

Add `"evaluator"` role to all users who need it (see `users` table). You can
use a loop like this:

```R
for (user_id in users$user_id) {
  access <- rbind(access, data.frame(
    user_id = user_id,
    deployment_id = deployments$deployment_id,
    user_roles = "evaluator"
  ))
}
db_write_table(con, "access", access, mode = "upsert")
```

The deployment materials repeat the materials for each deployment.
To do this part, we read in the materials (filter if deployments involve
a subset), and add the deployment ID. Finally, write the table to the database:

```R
materials <- db_read_table(con, "materials")

deployment_materials <- data.frame(
  deployment_material_id = paste0(
    rep(
      c("deployment1_", "deployment2_"),
      each = length(materials$material_id)
    ),
    materials$material_id
  ),
  deployment_id = rep(
    c("deployment1", "deployment2"),
    each = length(materials$material_id)
  ),
  material_id = materials$material_id
)

db_write_table(
  con,
  "deployment_materials",
  deployment_materials
)
```

Deployment subunits can be set up as a regular grid based on the extent
of a reference raster (distribution maps) and the spatial points 
(observations) object:

```R
# raster template
r <- read_file(file.path(path, "predictions", "LEYE_can71_2020.tif"))
# spatial points template
xy <- read_file(file.path(path, "data", "observations_can71.csv"))
xy <- xy[, c("lat", "lon")]
xy <- sf::st_as_sf(xy, coords = c("lon", "lat"))
sf::st_crs(xy) <- 4326

prep_deployment_subunits(
  deployment_id = "deployment1",
  x = NULL,
  reference_raster = r,
  reference_points = xy,
  n = c(20, 20),
  square = TRUE
)
```

Alternatively, deployment subunits can be defined based on custom vector layers (e.g., ecoregions/ecoprovinces). In this case, the subunit vector layer will be
transformed to epsg:4326 and clipped to the bounding box of the reference
layers:

```R
# ecoprovinces
su <- sf::st_read(file.path(path, "subunits", "ecoprovinces.shp"))
su$subunit_id <- su$ECOPROVINC

prep_deployment_subunits(
  deployment_id = "deployment2",
  x = su,
  reference_raster = r,
  reference_points = xy
)
```

## Local deployment

To share the deployment with evaluators, follow these steps.

First, test that your deployment works:

```R
  user_id <- "testuser"
  sdm_tool(user = user_id)
```

This should launch a pop-up window with the Shiny app. After 
confirming the deployment works, compress the deployment materials and the fresh database:

```R
od <- setwd("misc")
utils::zip(
  "./sdm_evaluation_results.zip",
  file.path(
    "sdm_evaluation_results",
    list.files("sdm_evaluation_results", recursive = TRUE)
  )
)
setwd(od)
```

This will create the `misc/sdm_evaluation_results.zip` file.
Send this file to your evaluators and ask them to follow the installation instructions in the evaluator's guide.

