---
title: SDM Evaluation Tool Report
output: pdf_document
params:
  user_id: "testuser"
  model_id: "bam_v5_can71"
  species_id: "BBWA"
  deployment_id: "deployment1"
  base_path: "../misc/test"
---

```{r include=FALSE}

#' Usage:
#' rmarkdown::render(
#'     input = "working/report-single-evaluation.Rmd", 
#'     params = list(
#'     user_id = "testuser",
#'     model_id = "bam_v5_can71",
#'     species_id = "BBWA",
#'     deployment_id = "deployment1",
#'     base_path = "../misc/test"
#' ))

library(sf)
library(terra)

# library(sdmEvalToolCore)
# library(sdmEvalToolUI)
devtools::load_all("../sdmEvalToolCore")
devtools::load_all("../sdmEvalToolUI")

sdmevaltool_options(base = params$base_path)

# values
user_id <- params$user_id
model_id <- params$model_id
species_id <- params$species_id
deployment_id <- params$deployment_id

# pull data
con <- db_connect()
t_dp <- db_read_table(con, "deployments") |>
    dplyr::filter(deployment_id == .env$deployment_id)
t_mo <- db_read_table(con, "models") |>
    dplyr::filter(model_id == .env$model_id)
t_sp <- db_read_table(con, "species") |>
    dplyr::filter(species_id == .env$species_id)
t_u <- db_read_user_info(con, user_id = user_id, deployment_id = deployment_id)
d <- db_read_evaluations(
    con,
    deployment_id = "deployment1",
    user_id = "testuser"
)
q <- prep_evaluations("testuser")
e <- prep_evaluations("testuser") |>
    tidyr::unnest("answers")
db_disconnect(con)

# get get the components and paths
cp <- sdmEvalToolCore::components |>
    dplyr::filter(type == "material", component != "app")

# functions
read_fun <- function(component) {
    f <- make_target_path(
        sdmEvalToolCore::components$path[
            sdmEvalToolCore::components == component
        ],
        list(model_id = model_id, species_id = species_id)
    )
    read_file(f)
}
print_evals <- function(ee) {
    for (i in 1:nrow(ee)) {
        cat("\n#### Question ", i, "\n\n")
        cat("**Question**: ", ee$label[i], "\n\n")
        v <- ee$response[i][[1]]
        if (length(v) == 1L) {
            cat("**Answer**: ", unlist(v), "\n\n")
            m <- NULL
        } else {
            if (nrow(v) == 1L) {
                cat("**Answer**: ", v$subunit, "\n\n")
                m <- NULL
            } else {
                cat("**Answer**:\n\n")
                print(knitr::kable(v))
                m <- su
                m$values <- NA_character_
                for (j in nrow(v):1) {
                    m$values[m$subunit_id %in% v$subunits[j][[1L]]] <- v$values[j]
                }
            }
        }
        if (!is.null(m)) {
            plot(m[,"values"], lwd=0.5)
        }
        cat("\n")
    }

}


# get evaluation time
Tmin <- min(d$evaluation_create_time)
Tmax <- max(d$evaluation_modify_time, na.rm = TRUE)
Tdiff <- difftime(Tmax, Tmin, units = "mins")
```

# Settings

- Evaluator: `r t_u$usernale` (<`r t_u$user_email`>, `r t_u$user_affiliation`) [ID: `r t_u$user_id`]
- Evaluation: start time `r Tmin`, end time `r Tmin` (duration: `r round(Tdiff, 1)` minutes)
- Model: `r t_mo$model_name`, `r t_mo$model_description` [ID: `r t_mo$model_id`]
- Species: `r t_sp$english_name` (_`r t_sp$scientific_name`_) [ID: `r t_sp$species_id`]
- Deployment: `r t_dp$deployment_description` [ID `r t_dp$deployment_id`]

# Predictions

```{r echo=FALSE}
su <- deployment_subunits_prep(deployment_id)
r <- read_fun("spatial_prediction")

plot(r[[1]], main="Estimate")
plot(su[[attr(su, "sf_column")]], add=TRUE, lwd=0.5, border="grey")
```

```{r echo=FALSE}
plot(r[[2]], main="Variation")
plot(su[[attr(su, "sf_column")]], add=TRUE, lwd=0.5, border="grey")
```

```{r echo=FALSE,results="asis"}
ee <- e[endsWith(e$material_id, "spatial_prediction"),] |> print_evals()
```

# Observations

```{r echo=FALSE}
# xy <- read_fun("observations")
obs <- obs_prep(
    model_id,
    species_id,
    output_type = "points")
rast <- obs_prep_raster(obs, r)

plot(r[[1]], alpha=0.5, main="Observations")
plot(rast[["absence"]], col = "blue", add=TRUE, legend = FALSE)
plot(rast[["presence"]], col = "red", add=TRUE, legend = FALSE)
plot(su[[attr(su, "sf_column")]], add=TRUE, lwd=0.5, border="grey")
```

```{r echo=FALSE,results="asis"}
ee <- e[endsWith(e$material_id, "observations"),] |> print_evals()
```

# Model Summary

```{r echo=FALSE,results="asis"}
x <- read_fun("model_summary")
knitr::kable(x)
```

```{r echo=FALSE,results="asis"}
ee <- e[endsWith(e$material_id, "model_summary"),] |> print_evals()
```

# Model Fit

```{r echo=FALSE,results="asis"}
x <- read_fun("model_fit")
knitr::kable(x)
```

```{r echo=FALSE,results="asis"}
ee <- e[endsWith(e$material_id, "model_fit"),] |> print_evals()
```

# Predictor Raster

```{r echo=FALSE}
# too many layers, unclear how to display
# x <- read_fun("predictor_raster")
```

```{r echo=FALSE,results="asis"}
ee <- e[endsWith(e$material_id, "predictor_raster"),] |> print_evals()
```

# Predictor Metadata

```{r echo=FALSE}
# too many columns, unclear how to display
# x <- read_fun("predictor_metadata")
```

```{r echo=FALSE,results="asis"}
ee <- e[endsWith(e$material_id, "predictor_metadata"),] |> print_evals()
```
